---
title: "Its tough to be a caribou."
description: |
  Tidy Tuesday caribou tracking data.
author:
  - name: David Fox
    url: {}
date: 08-01-2020
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse) 
library(janitor) 
library(vroom) 
library(hrbrthemes) 
library(gt) 
library(here)
library(plotly)
library(sf)
library(lubridate)
library(mapview)
library(widgetframe)
library(tmap)

theme_set(theme_ipsum())
```

This week's data is about caribou in British Columbia, Canada.

Data: (<https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-06-23/readme.md>)

```{r get data}
tuesdata <- tidytuesdayR::tt_load('2020-06-23')

#locations <- tuesdata$locations
individuals <-  tuesdata$individuals
```

There are 2 data sets, one on individuals and one on locations. What data is there about the individuals?

{[skimr](https://github.com/ropensci/skimr)} is a great package for data overview. As a pro tip, you can use use {skimr} and git diffs to quickly see what has changed with a data set.

```{r skim}

skimr::skim(individuals)
```

Looks like a lot of missing data. What's the sex ratio of collared individuals?

```{r}
individuals %>%
  tabyl(sex) %>%
  adorn_pct_formatting() %>% gt()
```

They only collared 4 males -- that is a very skewed sex ratio. Are male caribou tremendously hard to capture for radio collaring?

What are the deploy off types (one of the few complete variables)?

```{r}
individuals %>%
  tabyl(deploy_off_type) %>%
  adorn_pct_formatting() %>%
  gt()
```

So about 1/5 died, 1/3 have been removed, over 1/3 are unknown, and a bunch are 'other'. That is a lot of radio collars, which are not cheap, with an unknown status. Is there anything useful in the 'deploy_off_comments'?

```{r}
individuals %>%
  tabyl(deploy_off_comments) %>%
  adorn_pct_formatting() %>%
  gt()
```

Not really - more missing data- and what the heck does "Infometrics" mean.. do these researchers have a graduate adviser?[^1]

[^1]: I'm being snarky, but I spent 5 years collecting biological field data. One of the big challenges is taking the time to document your research and be a good data steward. There may be mitigating factors I'm not aware of with this data set, but if I was managing this project as a researcher or a grantor I might not be very happy with the quality of this data collection.

Being somewhat morbid, I'm curious about how these caribou died.

```{r}
individuals %>%
  filter(deploy_off_type == 'dead') %>%
  mutate(death_cause = str_to_lower(death_cause)) %>%
  tabyl(death_cause) %>%
  adorn_pct_formatting() %>%
  gt()
```

Let's do a bit of recoding on these comments.

```{r}
deceased <- individuals %>%
  mutate(death_cause = str_to_lower(death_cause)) %>% 
           filter(deploy_off_type == 'dead') %>% 
  mutate(cause_of_death = case_when(
    death_cause == 'accidental, collar still in field, inaccessible' ~ "accident",
    death_cause == 'accidental.  caught in tree well'                 ~ "accident",
    death_cause == 'predation - grizzly'                             ~ "predation - grizzly bear",
    death_cause == 'collar still in field'                           ~ "unknown",
    death_cause == 'unknown.  collar inaccessible and still in field.'~ "unknown",
    death_cause == 'unknown.  found 50m from trend open pit mine.  area littered with rock from blasts from the mine.' ~ "explosion",
    death_cause == 'unknown.  suspected predation' ~ "suspected predation",
    death_cause == 'unknown.  suspected wolf (or possibly wolverine) predation' ~ "suspected predation",
    death_cause == 'unknown.  suspected wolf predation' ~ "suspected predation",
    is.na((death_cause)) ~ 'unknown',
    TRUE ~ death_cause
))

deceased %>%
  tabyl(cause_of_death) %>%
  adorn_pct_formatting() %>%
  gt()
```

So these researchers need some QC on their data collection, but they're busy scientists so we'll forgive them. It looks like wolf predation is an issue for caribou -but also grizzly bears. One poor caribou was hit by a train, another by a car, and it seems like a researcher is suggesting -possibly- one got killed by debris from a blast at a mine! Being a caribou sounds rough.

Let's look at the locations of where these caribou met their fates. While ggplot spatial functions are evolving rapidly, The {[sf](https://r-spatial.github.io/sf/index.html)} package has become the default package for working with spatial data and plays nicely with tidyverse principals and tools. Just a note on coordinate precision. Each decimal place is about a factor of 10. These locations are accurate to within about 10 meters. For more on this -- XCD of course has the definitive [summary](https://xkcd.com/2170/)

```{r}
deceased_sf <- deceased %>%
  filter(!is.na(deploy_off_longitude)) %>%
  st_as_sf(., coords = c("deploy_off_longitude", "deploy_off_latitude"), crs = 4326)
```

{[tmap](https://github.com/mtennekes/tmap)} is an excellent package for interactive mapping in R that goes a bit beyond ggplot. Other good mapping packages worth checking out are {[Mapview](https://github.com/r-spatial/mapview)} and {[mapdeck](https://github.com/SymbolixAU/mapdeck)}.

```{r echo=TRUE, fig.height=8, fig.width=8, layout="l-body-outset"}
tmap_mode("view")

tm_shape(deceased_sf) +
  tm_dots(col = "cause_of_death", 
             palette = "plasma")

```
