---
title: "Space Above and Beyond"
description: |
  Tidy Tuesday Astronaut data.
author:
  - name: David Fox
    url: {}
date: 08-01-2020
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height = 4)

library(tidytuesdayR)
library(tidyverse)
library(hrbrthemes)
library(skimr)
library(janitor)
library(ggdist)
library(ggbeeswarm)
library(ggrepel)
library(gt)
library(plotly)

theme_set(theme_ft_rc())
```

#### This week's Tidy Tuesday data is about astronauts.

First I'll load the data and examine it with the eminently useful {skmir}.

```{r get data}
tuesdata <- tidytuesdayR::tt_load('2020-07-14')

astronauts = tuesdata$astronauts

skim(astronauts)

```

#### Roles in spaceflight

Occupation, which seems to mean mission role, looks like an interesting variable to explore. First order to is to eliminate any capitalization inconsistencies - which seem to plague all data. There is a lot of discussion about the pros and cons of SQL, but one of the advantages of a database structure is the ability to create rules that keep inconsistent data from ever getting into the data stream.

```{r}
astronauts %>%
  mutate(occupation = str_to_lower(occupation)) %>%   # always seem to be capitalization issues
  tabyl(occupation) %>%
  adorn_pct_formatting() %>% 
  gt()
```

I'm not sure what some of these roles are, also a little re-coding is in order to deal with redundant data categories. Some external research indicates that 'msp' probably stands for mission specialist, and 'psp' stands for payload specialist. This [wikipedia page](https://en.wikipedia.org/wiki/Astronaut_ranks_and_positions#Ranks) has more details. I'll recode these and lump all the tourists together -- sorry lone space journalist. One thing to watch out for with *case_when* is to make sure to use double quotes, I was using single quotes and could not figure out why things were not working.

```{r}
astronauts_clean <- astronauts %>% 
  select(name, sex, occupation, year_of_mission, hours_mission) %>% 
  mutate(occupation = str_to_lower(occupation)) %>%
  mutate(role = case_when(
    occupation == "msp" ~ "mission specialist",
    occupation == "psp" ~ "payload specialist",
    TRUE ~ occupation),
    role_factor = fct_lump_n(role, 5))
```

### Gender balance in space flight roles.

Some of the early pioneers in spaceflight were women. I'm curious about how the current gender roles look.

```{r}
astronauts_clean %>%
    tabyl(role_factor, sex) %>%
  adorn_totals("row") %>% 
  adorn_percentages("row") %>% 
  adorn_pct_formatting() %>%
  adorn_ns() %>%
  adorn_title("combined", row_name = "Role", col_name = "gender") %>% 
  gt()
```

Not great - only 11% of astronauts have been women. Only 1% of mission commanders have been women, and less than 4% have been pilots. How does this look over time?

```{r, layout="l-body-outset", fig.height=6}

set.seed(567)

p1 = astronauts_clean%>% 
  ggplot(aes(x= year_of_mission, y= sex, color = role_factor,
             text = paste(year_of_mission, "<br>",name,"<br>",role_factor))) +
  geom_quasirandom(alpha = .4,
                   size = 2,
                   groupOnX = FALSE) +
  labs(title = 'Gender roles in space flight',
       y = NULL, 
       x = NULL,
       color = "Role",
       caption = 'TidyTuesday 2020-07-14') +
  scale_color_viridis_d(option ="plasma") +
  theme(legend.position = "bottom")

ggplotly(p1, tooltip = "text")

```

[Valentina Tereshkova](https://en.wikipedia.org/wiki/Valentina_Tereshkova) was the first woman to travel to space - she did a solo flight in 1963 that lasted for 3 days. If you read her wikipedia page, she is clearly a bad ass. There was not another woman pilot in space until 1997 - [Eileen Collins](https://en.wikipedia.org/wiki/Eileen_Collins). It seems like the role of pilot has disappeared since the Space Shuttle was retired, but there are still male mission commanders, while the last woman commander was in 2007.

### Waxing and waning of space flight.

The previous chart shows a several strong pulses in the number of missions.

```{r, layout="l-body-outset", fig.height=6}
missions  = astronauts %>% 
  distinct(mission_title, .keep_all = TRUE) %>% 
  select(year_of_mission, mission_title, ascend_shuttle, in_orbit, hours_mission)

p2 = missions %>% group_by(year_of_mission) %>% 
  summarize(mission_count = n()) %>% 
  ggplot(aes(x= year_of_mission,
             y = mission_count,
             color = mission_count,
             size = mission_count,
             text = paste(year_of_mission, ":", mission_count))) +
    geom_point(alpha = .5) + 
    scale_color_viridis_c(option = "plasma") +
  labs(title = str_wrap("After a heyday in the 90s the number of flights has declined to the level of the 1960's."),
       x= NULL,
       y = 'missions') +
  theme(legend.position = "none")

ggplotly(p2, tooltip = "text")
```

### Mission hours have increased dramatically with space stations.

```{r fig.height=5, layout="l-body-outset", fig.height=6}
p3 = ggplot(missions, aes(x= year_of_mission,
                          y = hours_mission,
                          size = hours_mission,
                          color = hours_mission,
                          text = paste(in_orbit, "<br>", hours_mission))) +
  geom_jitter(alpha = .5) +
  #geom_smooth(se = FALSE) +
  scale_color_viridis_c(option = "plasma") +
  labs(title = "Although number of missions has declined, \ntime is space has increased dramitically.",
       x= "",
       y = "mission hours") + 
  theme(legend.position = 'none') +
  scale_y_sqrt()


ggplotly(p3, tooltip = "text")
```
